Evolution of precopulatory and postcopulatory inbreeding behaviour

### 2016-09-09 ms_pre_post.Rmd
Minor edits up to Results

<!--- TODO: OVERVIEW OF THE POINT OF THE MODELLING --->
Here we apply a model of female and male reproductive interactions and associated inbreeding depression in offspring viability to quantify evolution of three separate phenotypes including female tendency for polyandry, female pre-copulatory inbreeding strategy (mating strategy), and female post-copulatory inbreeding strategy (fertilisation strategy). We use individual-based modelling and *in silico* experiments to test the hypothesis that, given severe inbreeding depression in offspring viability, evolution of inbreeding avoidance through fertilisation bias causes selection for evolution of polyandry. Additionally, we test the hypothesis that costs associated with polyandry and inbreeding strategies will strongly affect evolution of each such that that inbreeding avoidance through mate choice but not fertilisation bias will evolve when inbreeding mating strategy is less costly than the combined cost of inbreeding fertilisation strategy and tendency for polyandry polyandry, while inbreeding avoidance through fertilisation bias but not mate choice will evolve when inbreeding mating strategy is more costly than the combined costs of inbreeding fertilisation strategy and tendency for polyandry. Finally, we test the hypothesis that constraints on evolution of one inbreeding strategy (e.g., inbreeding mating strategy) will increase the strength of selection on evolution of a complementary inbreeding strategy (e.g., inbreeding fertilisation strategy), and that evolution of one inbreeding strategy will be precluded if it is introduced into a population that is already fixed for a complementary inbreeding strategy that is adaptive. We thereby provide a framework for predicting evolution of pre-copulatory and post-copulatory inbreeding avoidance across reproductive systems with different relative costs and different magnitudes of inbreeding depression in offspring viability.

@@

We track interactions among discrete individuals in a small focal population that persists over multiple generations (Key individual traits, parameter values, and variables are described in Table 1). Individuals within this population are either female or male, and can be related to each other if they share a common ancestor. We record the coefficient of kinship ($k$), defined as the probability that a randomly selected homologous allele from each individual is identical-by-descent [@Lynch1998], between all pairs of individuals in each generation using the population pedigree [see @Boyce1983; @Duthie]. At the start of each generation, each female pays up to three independent costs of (1) tendency to engage in polyandry ($c_{P}$), (2) mating strategy with respect to inbreeding ($c_{M}$), and (3) fertilisation strategy with respect to inbreeding ($c_{F}$). The number of males that a focal female mates with ($N_{males}$) is determined by the effects of 20 alleles ($P_{a}$) across 10 diploid loci that combine additively to determine her genotype ($P_{g}$), and ultimately her phenotype for tendency to engage in polandry ($P_{p}$). A focal female's choice of mate(s) is affected by 20 alleles ($M_{a}$) across 10 diploid loci that combine additively to affect her genotype ($M_{g}$) and phenotype ($M_{p}$) for pre-copulatory inbreeding avoidance or preference, and kinship between the focal female and her potential mates. Similarly, after mating has occurred, post-copulatory fertilisation of a focal female's $n$ total offspring is affected by 20 alleles ($F_{a}$) across 10 diploid loci that combine additively to affect the genotype ($F_{g}$) and phenotype ($F_{p}$) of the degree to which fertilisation success is affected by kinship. After fertilisation, the viability of a focal female's $n$ offspring is decreased by inbreeding depression ($\beta$), and surviving offspring become adults. After inbreeding depression in offspring, $\rho$ adult male immigrants arrive in the population, then density effects reduce the entire population down to female ($K_{f}$) and male ($K_{m}$) carrying capacity. Over generations, we record mean values of alleles underlying polyandry, inbreeding mating strategy, and inbreeding fertilisation strategy.

@@

The value of $P_{p}$ is affected by the summed values of her 20 $P_{a}$ alleles. Inheritence of all alleles models a diploid genetic architecture in which 10 $P_{a}$ alleles are randomly sampled from each parent with no physical linkage. Values of $P_{a}$ can take any real number [continuum-of-alleles model; @Kimura1965; @Lande1976; @Reeve2000; @Bocedi2014]. The tendency for polyandry genotype $P_{g}$ is defined by the sum of all of a focal female's 20 $P_{a}$ values, which might be negative or positive. However, the tendency for polyandry phenotype $P_{p}$ is restricted to zero or positive values because the number of males that a female chooses to mate with cannot be negative. Hence, negative values of $P_{g}$ are translated to $P_{p}$ values of zero such that $P_{p} = 0$ when $P_{g} < 0$, but when $P_{g} \geq 0$, then $P_{p} = P_{g}$ (i.e., the phenotype value of polyandry equals the genotype value of polyandry if and only if the genotype value is greater than or equal to zero, else the phenotype value of polyandry always equals zero). The resulting $P_{p}$ is therefore modelled as a threshold trait [@Lynch1998; @Roff1996; @Roff1998], in which polyandry is expressed discretely at the threshold value of $P_{g} > 0$, but is determined by the consequence of continuous genetic variation arising from the effects of $P_{a}$ alleles that each contribute additively to a female's tendency toward polyandry. Negative genotypic values of $P_{g}$ therefore all have the same phenotype, resulting in consistently monandrous females. In contrast, positive values of $P_{g}$ and therefore $P_{p}$ can lead to different degrees of realised polyandry in the form of discrete numbers of mates ($N_{males}$).

@@

Each male within the population is a potential mate for any female to choose. We therefore assume that there are no opportunity costs associated with male mating such that mating with one female would reduce a male's ability to mate with and sire the offspring of any other female [e.g., @Waser1986]. While all males are therefore available as potential mates, we assume that a female mates with her total allotment of $N_{males}$ without replacement, meaning that $N_{males}$ models the total number of distinct males that a female chooses to mate with rather than the number of times that she mates. In the unlikely event that a female's $N_{males}$ exceeds the total number of males in the population, then she mates with all of the males in the populaton. Most often, $N_{males}$ will be a small subset of the available males in the population [@Duthie], so a focal female will be able to preferentially mate with males that are close kin (high $k$), distant kin (low $k$), or non-kin ($k=0$) through the expression of her inbreeding mating strategy phenotype ($M_{p}$). Like polyandry allele values, inbreeding mating strategy allele values ($M_{a}$) can take any real number. However, unlike polyandry, the sum of all of a female's 20 $M_{a}$ alleles equals both her inbreeding mating strategy genotype ($M_{g}$) and her inbreeding mating strategy phenotype ($M_{p}$). Negative or positive values of $M_{p}$ cause a female to avoid or prefer mating with kin, respectively. Values of $M_{p}=0$ causes females to mate randomly with respect to kinship.

@@

After offspring viability is realised, $\rho$ adult immigrants are added to the focal population. The coefficient of kinship between an immigrant and all other individuals (including native individuals and other immigrants) always equals zero ($k_{ij}=0$). Because kinship coefficient values are calculated directly from the pedigree of the focal population, the addition of immigrants thereby introduces non-kin into the population at a consistent rate and prevents the mean kinship coefficient of the population from asymptoting to a value of 1 over generations. If the focal population was instead closed ($\rho=0$), then all alleles would quickly be expected to become identical-by-descent among descendants of the ancestral population. To avoid having immigrants affect the evolution of tendency for polyandry ($P_{p}$), inbreeding mating strategy ($M_{p}$), and inbreeding fertilisation strategy ($F_{p}$), immigrants are always male and can therefore be chosen based on their kinship values of zero but not actively affect reproductive decisions through the expression of $P_{p}$, $M_{p}$, or $F_{p}$. To further avoid having immigrants affect evolutionary dynamics, allele values underlying tendency for polyandry ($P_{a}$), inbreeding mating strategy ($M_{a}$), and inbreeding fertilisation strategy ($F_{a}$) are randomly sampled from normal distributions with means and standard deviations of $P_{a}$, $M_{a}$, and $F_{a}$ values equal to the means and standard deviations of those in the focal population at the time of immigration. Overall, this model of immigration effectively assumes that the focal population receives immigrants from other nearby populations that are subject to the same selection on $P_{p}$, $M_{p}$, and $F_{p}$.

@@

To analyse our model, we track mean values of alleles underlying tendency for polyandry ($P_{a}$), inbreeding mating strategy ($M_{a}$), and inbreeding fertilisation strategy ($F_{a}$) in a focal population. We present evolutionary dynamics of these mean allele values across different parameter combinations test our four focal hypotheses (see below). We simulate up to eight possible permutations of values of costs of tendency for polyandry ($c_{P}$), inbreeding mating strategy ($c_{M}$), and inbreeding fertilisation strategy ($c_{F}$) equal to either $0$ or $0.02$ [@Duthie observed $0.02$ to be a strong cost of tendency for polyandry] for values of $\beta = 3$ ($8$ total combinations). In all simulations, we record mean values of $P_{a}$, $M_{a}$, $F_{a}$ in each generation. Each of combination of parameter values simulated is replicated $40$ times, and grand mean values and standard deviation of means is calculated in each generation across replicate simulations. These analyses allowed us to infer how allele values change over generations in response to costs, but also in response to the changing values of other alleles and therefore potentially interacting phenotypes (e.g., selection on inbreeding fertilisation strategy might be weak if polyandry is rare). For all replicates, we set the maximum number of generations to 10000, which exploratory simulations and previous modelling [@Duthie] showed to be a sufficient number of generations for inferring long-term dynamics of mean allele values and therefore selection on phenotypes. For all replicates, we also set the number of immigrants per generation to $\rho=5$, which produced a range of kin and non-kin in each generation such that females could express inbreeding strategies. Females always produced $n=8$ offspring, which was sufficient to keep populations consistently near carrying capacities and avoid population extinction. Carrying capacities for females ($K_{f}$) and males ($K_{m}$) were both set to 100 because previous modelling has shown that populations of this size are small enough that mating encounters between kin occur with sufficient frequency for selection on inbreeding strategy, but populations are also not so small that selection is typically overwhelmed by genetic drift (Duthie and Reid *in press*).

@@

We first test our hypothesis that females' ability to bias fertilisation with respect to kinship causes selection for polyandry by comparing simulations in which tendency for polyandry, inbreeding mating strategy, and inbreeding fertilisation strategy can all evolve to simulations in which inbreeding fertilisation strategy is fixed to random mating with respect to kinship, as present in Figures 1 and 2, respectively. Comparison of Figures 1 and 2 shows that evolution of inbreeding fertilisation strategy facilitates evolution of polyandry, particularly when costs of tendency for polyandry and inbreeding fertilisation strategy are zero. For example, in Figure 2A, tendency for polyandry allele values consistently become negative over generations ($P_{a}<0$; decreasing red line), causing populations to be monandrous. Females in these populations avoid mating with kin ($M_{a}<0$; decreasing blue line), and because these females sample mates with replacement, males that remain after their initial mate choice are more likely to be relatives, causing selection for monandry because females are most likely to succeed in avoiding inbreeding in their initial mate choice [see also @Duthie]. However, Figure 1A shows that when a focal female can also bias the probability that a particular mate sires her offspring, inbreeding fertilisation strategy evolves to inbreeding avoidance ($F_{a}<0$; decreasing black line). Consequently, the ability to bias fertilisation to avoid inbreeding causes polyandry to become adaptive, and positive values of tendency for polyandry alleles increase in frequency ($P_{a}>0$; increasing red line of Figure 1A). Although non-zero costs decrease net selection for both tendency for polyandry and inbreeding fertilisation strategy (see below), across all combinations of costs, tendency for polyandry allele values are more positive when inbreeding fertilisation strategy can also evolve. Evolution of post-copulatory inbreeding avoidance therefore greatly facilitates evolution of polyandry.

@@

\vspace{100 mm}

@@

>>>>

<!--- TODO: OVERVIEW OF THE POINT OF THE MODELLING --->
Here we apply a model of female and male reproductive interactions and associated inbreeding depression in offspring viability to quantify evolution of three separate phenotypes including female tendency for polyandry, female pre-copulatory inbreeding strategy (mating strategy), and female post-copulatory inbreeding strategy (fertilisation strategy). We use individual-based modelling and *in silico* experiments to test the hypothesis that, given severe inbreeding depression in offspring viability, evolution of inbreeding avoidance through fertilisation bias causes selection for evolution of polyandry. Additionally, we test the hypothesis that costs associated with polyandry and inbreeding strategies will strongly affect evolution of each such that that inbreeding avoidance through mate choice but not fertilisation bias will evolve when inbreeding mating strategy is less costly than the combined cost of inbreeding fertilisation strategy and tendency for polyandry, while inbreeding avoidance through fertilisation bias but not mate choice will evolve when inbreeding mating strategy is more costly than the combined costs of inbreeding fertilisation strategy and tendency for polyandry. Finally, we test the hypothesis that constraints on evolution of one inbreeding strategy (e.g., inbreeding mating strategy) will increase the strength of selection on evolution of a complementary inbreeding strategy (e.g., inbreeding fertilisation strategy), and that evolution of one inbreeding strategy will be precluded if it is introduced into a population that is already fixed for a complementary inbreeding strategy that is adaptive. We thereby provide a framework for predicting evolution of pre-copulatory and post-copulatory inbreeding avoidance across reproductive systems with different relative costs and strong inbreeding depression in offspring viability.

@@

We track interactions among discrete individuals in a small focal population that persists over multiple generations (key individual traits, parameter values, and variables are described in Table 1). Individuals within this population are either female or male, and can be related to each other if they share a common ancestor. We record the coefficient of kinship ($k$), defined as the probability that a randomly selected homologous allele from each individual is identical-by-descent [@Lynch1998], between all pairs of individuals in each generation using the population pedigree [see @Boyce1983; @Duthie]. At the start of each generation, each female pays up to three independent costs of (1) tendency to engage in polyandry ($c_{P}$), (2) mating strategy with respect to inbreeding ($c_{M}$), and (3) fertilisation strategy with respect to inbreeding ($c_{F}$). The number of males that a focal female mates with ($N_{males}$) is determined by the effects of 20 alleles ($P_{a}$) across 10 diploid loci that combine additively to determine her genotype ($P_{g}$), and ultimately her phenotype for tendency to engage in polandry ($P_{p}$). A focal female's choice of mate(s) is affected by 20 alleles ($M_{a}$) across 10 diploid loci that combine additively to affect her genotype ($M_{g}$) and phenotype ($M_{p}$) for pre-copulatory inbreeding avoidance or preference, and kinship between the focal female and her potential mates. Similarly, after mating has occurred, post-copulatory fertilisation of a focal female's $n$ total offspring is affected by 20 alleles ($F_{a}$) across 10 diploid loci that combine additively to affect the genotype ($F_{g}$) and phenotype ($F_{p}$) of the degree to which fertilisation success is affected by kinship. After fertilisation, the viability of a focal female's $n$ offspring is decreased by inbreeding depression ($\beta$), and surviving offspring become adults. After inbreeding depression in offspring, $\rho$ adult male immigrants arrive in the population, then density effects reduce the entire population down to female ($K_{f}$) and male ($K_{m}$) carrying capacity. Over generations, we record mean values of alleles underlying polyandry, inbreeding mating strategy, and inbreeding fertilisation strategy.

@@

The value of $P_{p}$ is affected by the summed values of her 20 $P_{a}$ alleles. Inheritence of all alleles models a diploid genetic architecture in which 10 $P_{a}$ alleles are randomly sampled from each parent with no physical linkage. Values of $P_{a}$ can take any real number [continuum-of-alleles model; @Kimura1965; @Lande1976; @Reeve2000; @Bocedi2014]. The tendency for polyandry genotype $P_{g}$ is defined by the sum of all of a focal female's 20 $P_{a}$ values, which might be negative or positive. However, the tendency for polyandry phenotype $P_{p}$ is restricted to zero or positive values because the number of males that a female chooses to mate with cannot be negative. Hence, negative values of $P_{g}$ are translated to $P_{p}$ values of zero such that $P_{p} = 0$ when $P_{g} < 0$, but when $P_{g} \geq 0$, then $P_{p} = P_{g}$ (i.e., the phenotype value of polyandry equals the genotype value of polyandry if and only if the genotype value is greater than or equal to zero, else the phenotype value of polyandry always equals zero). The resulting $P_{p}$ is therefore modelled as a threshold trait [@Lynch1998; @Roff1996; @Roff1998], in which polyandry is expressed discretely at the threshold value of $P_{g} > 0$, but is determined by the consequence of continuous genetic variation arising from the effects of $P_{a}$ alleles that each contribute additively to a female's tendency toward polyandry. Negative genotype values of $P_{g}$ therefore all have the same phenotype, resulting in consistently monandrous females. In contrast, positive values of $P_{g}$ and therefore $P_{p}$ can lead to different degrees of realised polyandry in the form of discrete numbers of mates ($N_{males}$).

@@

Each male within the population is a potential mate for any female to choose. We therefore assume that there are no opportunity costs associated with male mating such that mating with one female would reduce a male's ability to mate with and sire the offspring of any other female [e.g., @Waser1986]. While all males are therefore available as potential mates, we assume that a female mates with her total allotment of $N_{males}$ without replacement, meaning that $N_{males}$ models the total number of distinct males that a female chooses to mate with rather than the number of times that she mates. In the unlikely event that a female's $N_{males}$ exceeds the total number of males in the population, then she mates with all of the males in the populaton. Most often, $N_{males}$ will be a small subset of the available males in the population [@Duthie], so a focal female will be able to preferentially mate with males that are close kin (high $k$), distant kin (low $k$), or non-kin ($k=0$) through the expression of her inbreeding mating strategy phenotype ($M_{p}$). Like polyandry allele values, inbreeding mating strategy allele values ($M_{a}$) can take any real number. However, unlike polyandry, the sum of all of a female's 20 $M_{a}$ alleles equals both her inbreeding mating strategy genotype ($M_{g}$) and her inbreeding mating strategy phenotype ($M_{p}$). Negative or positive values of $M_{p}$ cause a female to avoid or prefer mating with kin, respectively. Values of $M_{p}=0$ cause females to mate randomly with respect to kinship.

@@

After offspring viability is realised, $\rho$ adult immigrants are added to the focal population. The coefficient of kinship between an immigrant and all other individuals (including native individuals and other immigrants) always equals zero ($k_{ij}=0$). Because kinship coefficient values are calculated directly from the pedigree of the focal population, the addition of immigrants thereby introduces non-kin into the population at a consistent rate and prevents the mean kinship coefficient of the population from asymptoting to a value of 1 over generations. If the focal population was instead closed ($\rho=0$), then all alleles would quickly be expected to become identical-by-descent among descendants of the ancestral population. To avoid having immigrants affect evolution of tendency for polyandry ($P_{p}$), inbreeding mating strategy ($M_{p}$), and inbreeding fertilisation strategy ($F_{p}$), immigrants are always male and can therefore be chosen based on their kinship values of zero but not actively affect reproductive decisions through the expression of $P_{p}$, $M_{p}$, or $F_{p}$. To further avoid having immigrants affect evolutionary dynamics, allele values underlying tendency for polyandry ($P_{a}$), inbreeding mating strategy ($M_{a}$), and inbreeding fertilisation strategy ($F_{a}$) are randomly sampled from normal distributions with means and standard deviations of $P_{a}$, $M_{a}$, and $F_{a}$ values equal to the means and standard deviations of those in the focal population at the time of immigration. Overall, this model of immigration effectively assumes that the focal population receives immigrants from other nearby populations that are subject to the same selection on $P_{p}$, $M_{p}$, and $F_{p}$.

@@

To analyse our model, we track mean values of alleles underlying tendency for polyandry ($P_{a}$), inbreeding mating strategy ($M_{a}$), and inbreeding fertilisation strategy ($F_{a}$) in a focal population. We present evolutionary dynamics of these mean allele values across different parameter combinations test our four focal hypotheses (see below). We simulate up to eight possible permutations of values of costs of tendency for polyandry ($c_{P}$), inbreeding mating strategy ($c_{M}$), and inbreeding fertilisation strategy ($c_{F}$) equal to either $0$ or $0.02$ [@Duthie observed $0.02$ to be a strong cost of tendency for polyandry] for values of $\beta = 3$ ($8$ total combinations). In all simulations, we record mean values of $P_{a}$, $M_{a}$, $F_{a}$ in each generation. Each combination of parameter values simulated is replicated $40$ times, and grand mean values and standard deviation of means is calculated in each generation across replicate simulations. These analyses allowed us to infer how allele values change over generations in response to costs, but also in response to the changing values of other alleles and therefore potentially interacting phenotypes (e.g., selection on inbreeding fertilisation strategy might be weak if polyandry is rare). For all replicates, we set the maximum number of generations to 10000, which exploratory simulations and previous modelling [@Duthie] showed to be a sufficient number of generations for inferring long-term dynamics of mean allele values and therefore selection on phenotypes. For all replicates, we also set the number of immigrants per generation to $\rho=5$, which produced a range of kin and non-kin in each generation such that females could express inbreeding strategies. Females always produced $n=8$ offspring, which was sufficient to keep populations consistently near carrying capacities and avoid population extinction. Carrying capacities for females ($K_{f}$) and males ($K_{m}$) were both set to 100 because previous modelling has shown that populations of this size are small enough that mate encounters between kin occur with sufficient frequency for selection on inbreeding strategy, but populations are also not so small that selection is typically overwhelmed by genetic drift (Duthie and Reid *in press*).

@@

We first test our hypothesis that females' ability to bias fertilisation with respect to kinship causes selection for polyandry by comparing simulations in which tendency for polyandry, inbreeding mating strategy, and inbreeding fertilisation strategy can all evolve to simulations in which inbreeding fertilisation strategy is fixed to random mating with respect to kinship, as presented in Figures 1 and 2, respectively. Comparison of Figures 1 and 2 shows that evolution of inbreeding fertilisation strategy facilitates evolution of polyandry, particularly when costs of tendency for polyandry and inbreeding fertilisation strategy are zero. For example, in Figure 2A, tendency for polyandry allele values consistently become negative over generations ($P_{a}<0$; decreasing red line), causing populations to be monandrous. Females in these populations avoid mating with kin ($M_{a}<0$; decreasing blue line), and because these females sample mates with replacement, males that remain after their initial mate choice are more likely to be relatives, causing selection for monandry because females are most likely to succeed in avoiding inbreeding in their initial mate choice [see also @Duthie]. However, Figure 1A shows that when a focal female can also bias the probability that a particular mate sires her offspring, inbreeding fertilisation strategy evolves to inbreeding avoidance ($F_{a}<0$; decreasing black line). Consequently, the ability to bias fertilisation to avoid inbreeding causes polyandry to become adaptive, and positive values of tendency for polyandry alleles increase in frequency ($P_{a}>0$; increasing red line of Figure 1A). Although non-zero costs decrease net selection for both tendency for polyandry and inbreeding fertilisation strategy (see below), across all combinations of costs, tendency for polyandry allele values are more positive when inbreeding fertilisation strategy can also evolve. Evolution of post-copulatory inbreeding avoidance therefore greatly facilitates evolution of polyandry.

@@

@@

\vspace{60 mm}
